---
title: "Project 3: Matrix Factorization methods"
author: "Ilya Kats"
date: "June 26, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      warning = FALSE, 
                      message = FALSE)
```

### Introduction

This project is based on the work done in Project 2 (https://rpubs.com/ilyakats/data643proj2). It adds SVD to explore recommender systems further. 

```{r}
# Required libraries
library(recommenderlab)
library(dplyr)
library(tidyr)
library(ggplot2)
```

### Data Set

The data set is courtesy of __MovieLens__ project and it was downloaded from https://grouplens.org/datasets/movielens/.

```{r}
# Data import
ratings <- read.csv(paste0("https://raw.githubusercontent.com/ilyakats/CUNY-DATA643/",
                           "master/Project%202/ml-latest-small/ratings.csv"))
titles <- read.csv(paste0("https://raw.githubusercontent.com/ilyakats/CUNY-DATA643/",
                          "master/Project%202/ml-latest-small/movies.csv"))

# Convert to matrix
movieMatrix <- ratings %>% 
  select(-timestamp) %>% 
  spread(movieId, rating)
row.names(movieMatrix) <- movieMatrix[,1]
movieMatrix <- as.matrix(movieMatrix[-c(1)])
movieRealMatrix <- as(movieMatrix, "realRatingMatrix")
```

Our movie matrix contains 671 users and 9,066 items/movies. 

### SVD: Singular Value Decomposition

Before performing SVD, we have one preparatory step. Since SVD works on real matrices only, we will assume that non-existing rating is the same as rating of 0 and replace all `NA` with 0.

We perform SVD using R's base function `svd`. 

```{r}
# Replace NAs with 0
movieMatrix[is.na(movieMatrix)] <- 0

# Perform SVD
movieSVD <- svd(movieMatrix)
```

This process generates 671 concepts. Clearly to be usable we need to reduce number of dimensions/concepts by setting some singular values in the diagonal matrix $\Sigma$ to 0. Per Leskovec (_Mining of Massive Datasets_, 2014, p. 424), we will retain enough singular values to make up 90% of the energy of $\Sigma$.

```{r}
n <- length(movieSVD$d)
total_energy <- sum(movieSVD$d^2)
for (i in (n-1):1) {
  energy <- sum(movieSVD$d[1:i]^2)
  if (energy/total_energy<0.9) {
    n_dims <- i+1
    break
  }
}
```

This process leaves us with 274 dimensions/concepts. This is still a high number, but much more manageable than 671 (almost 60% reduction).

### Summary

TBD